### 2018-03-30 paper.Rmd
started writing teaching in childes-db section

```{r load_packages, echo = FALSE, warning=FALSE}
library(papaja)
library(tidyverse)
library(childesr)
library(directlabels)
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

@@

The foundational resource for studies of language acquisition is recordings and transcripts of children. Starting with Roger Brown's [-@brown1973] foundational work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) has formed the basis for this field. But creating individual transcripts is extremely hard work. Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe. Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.

@@

At its core, childes-db is a database consisting of a set of linked tabular data stores where rows corresponding to records. The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word. Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods. Deviations from the citation form—reflecting the deviations from citation forms that are particularly common in the course of language development— are kept as a separate (possibly null) field associated with each token.

Other tables in the database comprise a hierarchical system of collections built on top of tokens—*utterance*, *transcription*, *corpus*, and *collection*—which stores attributes appropriate for each level of description. Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.  An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes. [[ How are we handling the long tail of tiers?]]. A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES. A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus. Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES. Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction.

@@

childes-db offers a collection of browser-based interactive visualizations that enable exploration of the data through easy-to-use interfaces that don’t require any programming knowledge. These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without the technical overhead. The visualizations available are the following: Frequency Counts (for lexical usage statistics), MLU (to chart syntactic growth), and Population Viewer (to get a better understanding of the size of various corpora in the database). All three of these visualizations were built using Shiny, an R package that enables easy building of web applications in R. For each visualization, a user specifies a subset of the data, in particular the collection, corpus, child (target child in the session), and speaker (mother, father, etc.). The user may also restrict the age range of the child with slider bars. Any number of corpora, children, and speakers may be selected, allowing for aggregate counts across a whole range of data, a task that could not have been done with ease using previous methods. Along with the visualizations, childes-db makes the raw data itself available for download under a separate “Table” tab; this allows researchers to further explore and analyze the data they see in the plots.

@@

The Frequency Counts app allows users to enter a word in a text box and see the number of times this word was uttered across age for a particular subset of the data (in parts per million), whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake. Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.

### MLU

MLU (mean length of utterance) is an important measure used to assess the syntactic complexity in children’s speech. The MLU app plots MLU curves across age for a given subset of data, again specified by collection, corpora, children, and speakers. Because of the underlying gaps in available CHILDES morphological data, MLUw (mean length of utterance in words) is computed and not MLUm (mean length of utterance in morphemes). In the English language, these two measures have been shown to be highly correlated. As with the Frequency Counts app, caregiver’s MLU can be plotted alongside children.

@@

color_counts <- get_types(role = c("Mother", "Father"), age = c(12,60), 
                              type = colors)

count_estimates <- color_counts %>%
  mutate(age = target_child_age / 30.5,
         age_bin = as.integer(floor(age / 6) * 6),
         color = tolower(gloss)) %>%
  group_by(age_bin, color, target_child_id) %>%
  summarise(count = sum(count)) %>%
  summarise(mean = mean(count, na.rm = T),
            sem = sd(count, na.rm = T) / sqrt(sum(!is.na(count)))) %>%
  left_join(graph_colors) %>%
  mutate(color = factor(color, levels = colors))

ggplot(count_estimates, aes(x = age_bin, y = mean, ymin = mean - sem, 
                            ymax = mean + sem, color = color, label = color)) + 
  geom_pointrange(position = position_dodge(.5)) + 
  geom_smooth(se = F) +
  facet_wrap(~ color) + 
  theme_classic(base_size = 14) +
  scale_color_manual(values = graph_colors$graph_color) +
  scale_x_continuous(breaks = seq(12, 60, 12),
                     name = "age (months)") + 
  geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) + 
  theme(legend.position = "none")
```

@@

## Child-language class tutorials
## Data-science tutorials

>>>>

```{r load_packages, echo = FALSE, warning=FALSE}
library(papaja)
library(tidyverse)
library(childesr)
library(directlabels)
library(koRpus)
knitr::opts_chunk$set(cache = TRUE, echo = FALSE)
```

@@

The foundational resource for studies of language acquisition is recordings and transcripts of children. Starting with Roger Brown's [-@brown1973] foundational work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field. But creating individual transcripts is extremely hard work. Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe. Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.

@@

At its core, childes-db is a database consisting of a set of linked tabular data stores where rows correspond to records. The smallest unit of abstraction tracked by the database is a token, treated here as the citation orthographic form of a word. Using the standardized written form of the word facilitates the computation of lexical frequency statistics, for comparison or aggregation across children or time periods. Deviations from the citation form -- reflecting the deviations from citation forms that are particularly common in the course of language development -- are kept as a separate (possibly null) field associated with each token.

Other tables in the database comprise a hierarchical system of collections built on top of tokens -- *utterance*, *transcription*, *corpus*, and *collection* -- which stores attributes appropriate for each level of description. Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.  An *utterance* contains one or more words and includes fields such as the utterance type (declarative, question, etc.), total number of tokens, and the total number one morphemes. [[ How are we handling the long tail of tiers?]]. A *transcription* consists of one or more utterances and includes the date collected, the name of the target child and age in days (if defined), and the filename from CHILDES. A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown or Providence corpus. Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES. Because every entity can be linked to a top-level collection, childes-db places content from all languages in the same table for a given level of abstraction.

@@

childes-db offers a collection of browser-based interactive visualizations that enable data exploration through easy-to-use interfaces that don’t require any programming knowledge. These tools are intended for casual users who seek to extract useful developmental indices from CHILDES without any technical overhead. The visualizations available through childes-db are the following: Frequency Counts (lexical statistics), Derived Measures (lexical diversity), and Population Viewer (corpora statistics). All three of these visualizations were built using Shiny, an R package that enables easy building of web applications in R. For each visualization, a user specifies a subset of the data (in particular the collection, corpus, child, and speaker) and is also able to restrict the age range of the child with slider bars. Any number of corpora, children, or speakers may be selected for analysis, allowing for aggregate counts across a whole range of data in the CHILDES system, a task that could not have been easily done using previous methods. Along with the visualizations, childes-db makes the raw data itself available for download under a separate “Table” tab; this allows researchers to further explore the underlying data from the plots.

@@

The Frequency Counts app allows users to enter a word (or words) in a text box and see the frequency (in parts per million) in which this word was uttered across a child's age for a particular subset of the data, whether it is an individual child in the Brown corpus or all Spanish speaking children and their mothers. The ability to view the caregiver’s usage of a term alongside a child’s assists researchers in better understanding the relationship between input and uptake. Users can also plot word trajectories alongside each other (such as “mom” and “dad” or “front” and “back”) for convenient cross-word comparisons.

### Derived Measures

There are a number of ways to measure the lexical diversity of an individual's productive speech. Some of these measures, including MLU-w (mean length of utterance in words), TTR (type-token ratio), MTLD (measure of textual lexical diversity), and voc-d are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers. Because of the underlying gaps in available CHILDES morphological data, MLU-w is computed and not MLU-m (mean length of utterance in morphemes). In the English language, these two measures have been shown to be highly correlated. As with the Frequency Counts app, caregiver’s lexical diversity measures can be plotted alongside children's.

@@

color_counts <- get_types(collection = "Eng-NA", role = c("Mother", "Father"),
                          age = c(12,60), type = colors)

parents <- color_counts %>%
  distinct(collection_id, corpus_id, transcript_id, speaker_id) %>%
  left_join(get_speaker_statistics(collection = "Eng-NA")) %>%
  select(collection_id, corpus_id, transcript_id, speaker_id, num_tokens)

count_estimates <- color_counts %>%
  left_join(parents) %>%
  mutate(age = target_child_age / 30.5,
         age_bin = as.integer(floor(age / 6) * 6),
         color = tolower(gloss)) %>%
  group_by(age_bin, color, transcript_id) %>%
  summarise(count = sum(count), tokens = sum(num_tokens)) %>%
  mutate(prop = count/tokens * 1000000) %>%
  tidyboot::tidyboot_mean(prop) %>%
  left_join(graph_colors) %>%
  mutate(color = factor(color, levels = colors))

ggplot(count_estimates, aes(x = age_bin, y = empirical_stat, 
                            ymin = ci_lower, ymax = ci_upper,
                            color = color, label = color)) + 
  geom_pointrange(position = position_dodge(.5)) + 
  geom_point() +
  geom_smooth(method = "loess", se = F, span = .5) +
  facet_wrap(~ color, scales = "free_y") + 
  theme_classic(base_size = 14) +
  scale_color_manual(values = graph_colors$graph_color) +
  scale_x_continuous(breaks = seq(12, 60, 12),
                     name = "Age (months)") + 
  scale_y_continuous(name = "Frequency (per million words)") +
  geom_dl(method = list(dl.trans(x=x +.2), "last.qp", cex=1)) + 
  theme(legend.position = "none")
```

@@

## In class demonstrations

Teachers of courses on early language acquisition often want to demonstrate interesting changes in children’s early productive language use as well as communicate the richness of information available in children's language environments. One method is to present static displays on slides or handouts that show either (a) text from parent-child conversations extracted from CHILDES or (b) data visualizations of changes in a variety of metrics related to production and/or input (e.g., MLU or Frequency). The challenge with static displays is that they cannot be modified during the course of teaching and they rely on the instructor predicting what examples will be compelling to their students.

Interactive, in-class demonstrations offer a solution, and can be a powerful way to explain complex concepts. One of our goals in creating the browser-based interactive visualization tools in childes-db is to provide instructors with a new tool for communicating what can be learned by analyzing children's language production and input. In this section, we walk through one example of using childes-db to teach a key concept in first language acquisition: [todo: XXX]. We also show how an instructor might make the in-class demonstration interactive via eliciting hypotheses from the class that are operationalized as queries to the interactive visualization tools.

Potential topics: concrete vs. abstract nouns, development of question words, uncertainty terms, verbs vs. nouns vs. function words.

Along with the interactive data visualizations, childes-db makes the raw data itself available for download under a separate "Table" tab. This allows researchers to further explore and analyze the data they see in the plots. This functionality provides a link to the second teaching application for childes-db: making interesting data accessible for tutorials and take-home programming assignments in more computationally-focused language courses. In the next section, we walk through one example of how an instructor could use childes-db to create a tutorial that scaffolds students' ability to ask interesting computational questions using child language data.

## Tutorials and assignments

[todo: pick a topic and write some r code with gaps for the students to fill in]
